Assessing the potential of Jacobian-free Newton-Krylov methods for cell-centred finite volume solid mechanics

### 2024-09-19 article.tex
Merge ivan branch changes

\paragraph{Case 2: Out-of-plane bending of an elliptic plate}
This 3-D, static, linear elastic test case (Figure \ref{fig:elliptic_plate}) consists of a thick elliptic plate (0.6 m thick) with a centred elliptic hole, with the inner and outer ellipses given as
\begin{eqnarray}
	\left(\frac{x}{2}\right)^2 + \left(\frac{y}{1}\right)^2 = 1 & \text{inner ellipse} \\
	\left(\frac{x}{3.25}\right)^2 + \left(\frac{y}{2.75}\right)^2 = 1 & \text{outer ellipse}
\end{eqnarray}
The case has been described by the National Agency for Finite Element Methods and Standards (NAFEMS) \cite{Hitchings1987}, and analysed using finite volume procedures by \citet{Demirdzic1997a} and \citet{Cardiff2016a}.
Symmetry allows one-quarter of the geometry to be simulated.
A constant pressure of 1 MPa is applied to the upper surface, and the outer surface is fully clamped.
The mechanical properties are: $E = 210$ GPa, $\nu = 0.3$.
Six successively refined hexahedral meshes are used, with cell counts of 45, 472 (Figure \ref{fig:elliptic_plate}), $4\,140$, $34\,968$, $287\,280$ and $2\,438\,242$.
\begin{figure}[htbp]
   \centering
%\includegraphics[width=0.5\textwidth]{figures/elliptic_plate.pdf} 
	\subfigure[Mesh containing 472 hexahedral cells]
	{
		\includegraphics[width=0.45\textwidth]{figures/elliptic_plate.pdf} 
	}
	\subfigure[The equivalent (von Mises) stress distribution for the mesh with $2\,438\,242$ cells. The deformation is scaled by $1\,000$ and a translucent line indicates the outline of the undeformed geometry]
	{
		\includegraphics[width=0.45\textwidth]{figures/elliptic_plate_sigmaEq}  
	}
   \caption{Elliptic Plate case}
   \label{fig:elliptic_plate}
\end{figure}

@@

\paragraph{Case 3: Narrow T-section component under tension}
This case, proposed by \citet{Demirdzic1997a}, consists of a narrow engineering component with a T cross-section (Figure \ref{fig:narrowTmember}).
The case is 3-D, static, with linear elastic material behaviour.
Symmetry allows one-quarter of the geometry to be simulated.
A constant negative pressure of 1 MPa is applied to the lower surface, and the upper left surface is fully clamped.
The Young’s modulus is $E = 210$ GPa and Poisson’s ratio is $\nu = 0.3$.
A radius $5$ mm hole is located at the expected stress concentration.
\hl{Comment on the meshes}
\begin{figure}[htbp]
   \centering
%   \includegraphics[width=0.2\textwidth]{figures/narrowTmember.pdf} 
   \includegraphics[width=0.2\textwidth]{figures/placeholder.pdf} 
   \caption{Inflation of an idealised ventricle case geometry, mesh and loading conditions}
   \label{fig:narrowTmember}
\end{figure}

\paragraph{Case 4: Inflation of an idealised ventricle}
Inflation of an idealised ventricle (Figure \ref{fig:ventricle}) was proposed by \citet{Land2015} as a benchmark problem for cardiac mechanics software.
The case is 3-D, static, with finite hyperelastic strains.
The initial geometry is defined as a truncated ellipsoid:
\begin{eqnarray}
	x = r_s \sin(u) \cos(v), \quad
	y = r_s \sin(u) \sin(v), \quad
	z = r_l \cos(u)
\end{eqnarray}
where on the inner (endocardial) surface $r_s =7$ mm, $r_l = 17$ mm, $u \in \left[-\pi, -\arccos \left( \frac{5}{17} \right) \right]$ and $v \in \left[-\pi, \pi \right]$, while on the outer (epicardial) surface $r_s =10$ mm, $r_l = 20$ mm, $u \in \left[-\pi, -\arccos \left( \frac{5}{20} \right) \right]$ and $v \in \left[-\pi, \pi \right]$.
The base plane $z = 5$ mm is implicitly defined by the ranges for $u$.
%Constitutive parameters: isotropic, 
The hyperelastic material behaviour is described by the transversely isotropic constitutive proposed by \citet{Guccione} law, where the parameters $C = 10$ kPa, $b_f = b_t = b_{fs} = 1$.
The benchmark specifies an incompressible material; in the current work, incompressibility is enforced weakly using a penalty approach, where the bulk modulus parameter is chosen to be two orders of magnitude greater ($\kappa = 1000$ kPa) than the greatest shear modulus parameter.
The chosen parameters produce isotropic behaviour.
A pressure of 10 kPa is applied to the inner surface, and the base plane is fixed.
The geometry is meshed using a structured approach and is predominantly composed of hexahedra, with prism cells forming the apex.
The OpenFOAM utilities \texttt{blockMesh} and \texttt{extrudeMesh} are used to create the meshes.
Four successively refined meshes are examined: 1,620 (shown in Figure \ref{fig:ventricle}(a)), 12,960, 103,680, and 829,440 cells.
%Figure \ref{fig:ventricle} shows the mesh with 1,620 cells (1,512 hexahedra and 108 prisms).
\begin{figure}[htbp]
   \centering
%   \includegraphics[width=0.4\textwidth]{figures/ventricle} 
	\subfigure[Ventricle undeformed geometry, showing the mesh with 1,620 cells]
	{
		\includegraphics[height=0.46\textwidth]{figures/ventricle} 
	}
	\subfigure[Cross-section showing the equivalent (von Mises) stress distribution for the mesh with 829,440 cells]
	{
		\includegraphics[height=0.48\textwidth]{figures/ventricle_sigmaEq}  
	}
   \caption{Idealised ventricle case}
   \label{fig:ventricle}
\end{figure}

@@

\paragraph{Case 5: Cook's membrane}
Cook's membrane (Figure \ref{fig:cooks_membrane}) is a well-known bending-dominated benchmark case used in linear and non-linear analysis.
The current work considers the finite strain elastoplastic version \citep{Simo1992}. \hl{We could also consider the small-strain elastic version or hyperelastoplastic version}
The 2-D tapered panel (trapezoid) is fixed on one side and subjected to uniform shear traction on the opposite side.
The prescribed shear traction is $\tau = 0.3125$ MPa.
The vertices of the trapezoid (in mm) are (0, 0), (48, 44), (48, 60),  and (0, 44).
There are no body forces, and the problem is solved as statically as 2-D plane strain.
The problem is solved as static, using 30 equally-sized loading increments.
\begin{figure}[htbp]
   \centering
   \includegraphics[width=0.4\textwidth]{figures/cooksMembrane-geometry.png} 
   \caption{Cook's membrane case geometry (dimensions in mm) and loading conditions (fixed left, traction $\tau$ applied to right)}
   \label{fig:cooks_membrane}
\end{figure}

@@

%%%--------------------------------------------------------------------------------------------------------------------%%
%\subsection{Order of Accuracy Verification - Spherical Cavity}
%%%--------------------------------------------------------------------------------------------------------------------%%
%This section assesses the order of accuracy of the discretisation using the spherical cavity case (case 1).
%A secondary purpose is to demonstrate that the choice of solution algorithm (Jacobian-free Newton-Krylov vs segregated) does not affect the predictions, assuming iteration errors are small.
%
%The predicted axial ($\sigma_{zz}$) stress distribution is shown in Figure \ref{fig:spherical_cavity_stress}, where the peak stress ($2.xxx$ MPa) is just over twice the applied stress ($1$ MPa).
%This is in contrast to the classic 2-D circular hole in an infinite plate case, which exhibits a stress concentration factor of 3.
%A further difference with the 2-D circular hole case is that the local stresses more quickly return to the distant stress state as $R$ increases, with the local stresses being proportional to $\frac{1}{R^3}$ in the 3-D case and  $\frac{1}{R^2}$ in the 2-D case.
%\begin{figure}[htbp]
%   \centering
%%   \includegraphics[width=0.2\textwidth]{figures/spherical_cavity_stress.pdf} 
%   \includegraphics[width=0.2\textwidth]{figures/placeholder.pdf} 
%   \caption{Spherical cavity case: $\sigma_{zz}$ stress distribution \hl{use Reflect in ParaView to make a nice image}}
%   \label{fig:spherical_cavity_stress}
%\end{figure}
%
%Figure \ref{fig:spherical_cavity_accuracy} shows the discretisation errors ($L_2$ and $L_\infty$) in $\sigma_{zz}$ as a function of the average cell width.
%The errors can be seen to \hl{describe on the average/max error trends in graphs}.
%\hl{Comment on the order of accuracy: does it show what was expected or not?}
%\hl{Conclude 1: order of accuracy of the discretisation is XXX}
%\hl{Conclude 2: solution algorithm does not affect the answer}
%\hl{There is no need to show segregated results if they are the same: just state that they are the same}
%\hl{There is no need to show time and memory requirements: leave to the next section}
%\begin{figure}[htbp]
%   \centering
%%   \includegraphics[width=0.2\textwidth]{figures/spherical_cavity_error.pdf} 
%   \includegraphics[width=0.2\textwidth]{figures/placeholder.pdf} 
%   \caption{Spherical cavity case: discretisation error as a function of average cell width \hl{disp and stress errors if we have them}}
%   \label{fig:spherical_cavity_accuracy}
%\end{figure}

>>>>

\paragraph{Case 2: Out-of-plane bending of an elliptic plate}
This 3-D, static, linear elastic test case (Figure \ref{fig:elliptic_plate}) consists of a thick elliptic plate (0.6 m thick) with a centred elliptic hole, with the inner and outer ellipses given as
\begin{eqnarray}
	\left(\frac{x}{2}\right)^2 + \left(\frac{y}{1}\right)^2 = 1 & \text{inner ellipse} \\
	\left(\frac{x}{3.25}\right)^2 + \left(\frac{y}{2.75}\right)^2 = 1 & \text{outer ellipse}
\end{eqnarray}
The case has been described by the National Agency for Finite Element Methods and Standards (NAFEMS) \cite{Hitchings1987}, and analysed using finite volume procedures by \citet{Demirdzic1997a} and \citet{Cardiff2016a}.
Symmetry allows one-quarter of the geometry to be simulated.
A constant pressure of 1 MPa is applied to the upper surface, and the outer surface is fully clamped.
The mechanical properties are: $E = 210$ GPa, $\nu = 0.3$.
Six successively refined hexahedral meshes are used, with cell counts of 45, 472 (Figure \ref{fig:elliptic_plate}), $4\,140$, $34\,968$, $287\,280$ and $2\,438\,242$.
\begin{figure}[htbp]
   \centering
%\includegraphics[width=0.5\textwidth]{figures/elliptic_plate.pdf} ellipticPlate-geometry
	\subfigure[Mesh containing 472 hexahedral cells]
	{
		\includegraphics[width=0.45\textwidth]{figures/elliptic_plate.pdf} 
	}
	\subfigure[The equivalent (von Mises) stress distribution for the mesh with $2\,438\,242$ cells. The deformation is scaled by $1\,000$ and a translucent line indicates the outline of the undeformed geometry]
	{
		\includegraphics[width=0.45\textwidth]{figures/elliptic_plate_sigmaEq}  
	}
   \caption{Elliptic Plate case}
   \label{fig:elliptic_plate}
\end{figure}

@@

\paragraph{Case 3: Narrow T-section component under tension}
This case, proposed by \citet{Demirdzic1997a}, consists of a narrow engineering component with a T cross-section (Figure \ref{fig:narrowTmember}).
The case is 3-D, static, with linear elastic material behaviour.
Symmetry allows one-quarter of the geometry to be simulated.
A constant negative pressure of 1 MPa is applied to the lower surface, and the upper left surface is fully clamped.
The Young’s modulus is $E = 210$ GPa and Poisson’s ratio is $\nu = 0.3$.
A radius $5$ mm hole is located at the expected stress concentration.
\hl{Comment on the meshes}
%\begin{figure}[htbp]
%   \centering
%   \includegraphics[width=0.4\textwidth]{figures/narrowTmember-geometry.pdf} 
%   \caption{Inflation of an idealised ventricle case geometry, mesh and loading conditions}
%   \label{fig:narrowTmember}
%\end{figure}
\\
\hl{Philip, so far I have been doing pictures in xfig which is a bit outdated so I am switching to inkscape. I know you have this image of geometry from your article, so feel free to replace it. I made it purely to practice inkscaping.}
\begin{figure}[H]
	\centering
	\subfigure[Case geometry (considered quarter geometry)]
	{
		\label{fig:narrowTmember_geometry}
   		\includegraphics[scale=1]{figures/narrowTmember-geometry} 
   	}
   	\qquad
	\subfigure[Mesh with 1,620 cells]
	{
		\label{fig:narrowTmember_mesh}
   		\includegraphics[height=0.4\textwidth]{figures/placeholder.pdf}  
   	}
	\caption{Narrow T-section component under tension.}
	\label{fig:narrowTmember}
\end{figure}

\paragraph{Case 4: Inflation of an idealised ventricle}
Inflation of an idealised ventricle (Figure \ref{fig:ventricle}) was proposed by \citet{Land2015} as a benchmark problem for cardiac mechanics software.
The case is 3-D, static, with finite hyperelastic strains.
The initial geometry is defined as a truncated ellipsoid:
\begin{eqnarray}
	x = r_s \sin(u) \cos(v), \quad
	y = r_s \sin(u) \sin(v), \quad
	z = r_l \cos(u)
\end{eqnarray}
where on the inner (endocardial) surface $r_s =7$ mm, $r_l = 17$ mm, $u \in \left[-\pi, -\arccos \left( \frac{5}{17} \right) \right]$ and $v \in \left[-\pi, \pi \right]$, while on the outer (epicardial) surface $r_s =10$ mm, $r_l = 20$ mm, $u \in \left[-\pi, -\arccos \left( \frac{5}{20} \right) \right]$ and $v \in \left[-\pi, \pi \right]$.
The base plane $z = 5$ mm is implicitly defined by the ranges for $u$.
%Constitutive parameters: isotropic, 
The hyperelastic material behaviour is described by the transversely isotropic constitutive proposed by \citet{Guccione1995} law, where the parameters $C = 10$ kPa, $b_f = b_t = b_{fs} = 1$.
The benchmark specifies an incompressible material; in the current work, incompressibility is enforced weakly using a penalty approach, where the bulk modulus parameter is chosen to be two orders of magnitude greater ($\kappa = 1000$ kPa) than the greatest shear modulus parameter.
The chosen parameters produce isotropic behaviour.
A pressure of 10 kPa is applied to the inner surface, and the base plane is fixed.
The geometry is meshed using a structured approach and is predominantly composed of hexahedra, with prism cells forming the apex.
The OpenFOAM utilities \texttt{blockMesh} and \texttt{extrudeMesh} are used to create the meshes.
Four successively refined meshes are examined: 1,620 (shown in Figure \ref{fig:ventricle}(a)), 12,960, 103,680, and 829,440 cells.
%Figure \ref{fig:ventricle} shows the mesh with 1,620 cells (1,512 hexahedra and 108 prisms).
\begin{figure}[htbp]
   \centering
%   \includegraphics[width=0.4\textwidth]{figures/ventricle} 
	\subfigure[Ventricle undeformed geometry, showing the mesh with 1,620 cells]
	{
		\includegraphics[height=0.46\textwidth]{figures/ventricle} 
	}
	\subfigure[Cross-section showing the equivalent (von Mises) stress distribution for the mesh with 829,440 cells]
	{
		\includegraphics[height=0.48\textwidth]{figures/ventricle_sigmaEq}  
	}
   \caption{Idealised ventricle case}
   \label{fig:ventricle}
\end{figure}

@@

\paragraph{Case 5: Cook's membrane}
Cook's membrane (Figure \ref{fig:cooks_membrane}) is a well-known bending-dominated benchmark case used in linear and non-linear analysis.
The current work considers the finite strain elastoplastic version \citep{Simo1992}. \hl{We could also consider the small-strain elastic version or hyperelastoplastic version}
The 2-D tapered panel (trapezoid) is fixed on one side and subjected to uniform shear traction on the opposite side.
The prescribed shear traction is $\tau = 0.3125$ MPa.
The vertices of the trapezoid (in mm) are (0, 0), (48, 44), (48, 60),  and (0, 44).
There are no body forces, and the problem is solved as statically as 2-D plane strain.
The problem is solved as static, using 30 equally-sized loading increments.
%\begin{figure}[htbp]
%   \centering
%   \includegraphics[width=0.4\textwidth]{figures/cooksMembrane-geometry} 
%   \caption{Cook's membrane case geometry (dimensions in mm) and loading conditions (fixed left, traction $\tau$ applied to right)}
%   \label{fig:cooks_membrane}
%\end{figure}
\begin{figure}[H]
	\centering
	\subfigure[Case geometry and dimensions]
	{
		\label{fig:cooks_membrane_geometry}
   		\includegraphics[scale=1]{figures/cooksMembrane-geometry} 
   	}
   	\qquad
	\subfigure[Hexahedral mesh with 144 cells]
	{
		\label{fig:cooks_membrane_mesh}
   		\includegraphics[scale=0.14]{figures/cooksMembrane-mesh}  
   	}
	\caption{Cook's membrane.}
   \label{fig:cooks_membrane}
\end{figure}

@@
